---
title: "Prelim data analysis 2"
format: html
editor: visual
---

# Data Analysis 2 - Logistic Regression

In this file I step through the process of running a logistic regression using data from the file: Updated_Moose_Tree_Data.csv

The question that I'm asking is: Is there a relationship between whether a tree was interacted with and the tree species, dbh, and/or type of interaction?

## Set up

Prep the document

```{r}
rm(list=ls())
library(here)
library(tidyverse)
library(ggfortify)
```

Read in the data

```{r}
Tree <- read.csv(here("Data", "Updated_Moose_Tree_Data.csv"))
```

## Variables

The response variable is interaction, which is binary. 0 is no interaction, 1 is yes interaction.

The predictor variables are:

-   species (factor)

-   dbh_cm (ratio

-   Interaction.name (factor - need to change though, listed as character currently, even though I thought I changed it in the data exploration...)

```{r}
Tree$Interaction.name <- as.factor(Tree$Interaction.name)
```

I ran code in order to get the Interaction and Interaction.name variables in the data exploration, but I'll put it here as text as a reference.

*`pivot_longer` function used*

tree.updated2 \<- tree \|\> pivot_longer(cols = c("browsed", "stripped", "broken"), names_to = "Interaction.name", values_to = "Interaction")

*Need to make Interaction.name factor data*

tree.updated2$Interaction.name <- as.factor(tree.updated2$Interaction.name)

tree.updated2$Interaction.name <- factor(tree.updated2$Interaction.name)

*This will let me use the type of interaction as another predictor.*

Here's a [Link](https://www.datacamp.com/tutorial/logistic-regression-R) to a source I found on how to run logistic regression.

The model is a bit confusing, and needs information about "mixture" (either lasso model or ridge regression) and "penalty" (strength of regularization). There is also options for set_engine() and set_mode(), and fit(). I will await further instructions/assistance as this was as far as I could understand with the webpage.
