---
title: "Preliminary Data Analysis 1"
format: html
editor: visual
---

# Data Analysis 1 - Multiple regression

In this document I will go through the workflow of conducting a mutliple regression statistical analysis on the data from the file Updated_Moose_Plot_Data.csv

The question I'm asking is: Is there a relationship between the predictor variables and moose browse index?

## Set up

First I'll set up the document, then read in the data.

```{r}
rm(list=ls())
library(here)
library(tidyverse)
library(ggfortify)
```

```{r}
Plot <- read.csv(here("Data", "Updated_Moose_Plot_Data.csv"))
```

## Variables

The response variable is:

-   browse_index

The predictor variables I want to use are:

-   elev_m

-   dist_h2o_m

-   harvest_intensity_m2.ha

-   dist_dev_m

-   dist_conifer

-   deer_density

If binary data works well, I could also include these below, but I'll try without first so that I'm using all ratio data.

-   habitat type

    -   hilltop

    -   swamp

    -   wetland_forest

-   harvest

## Model

I am using this [Link](https://www.datacamp.com/tutorial/multiple-linear-regression-r-tutorial) to figure out the code for a multiple regression.

Based on that source, the model is fit similar to a linear regression, just with multiple predictors separated by +

```{r}
browseindex.model <- lm(browse_index ~ elev_m + dist_h2o_m + dist_conifer_m + harvest_intensity_m2.ha + dist_dev_m + deer_density, data = Plot)
```

## Check assumptions

The source recommends displaying the model residuals on a histogram now.

```{r}
model_residuals = browseindex.model$residuals

hist(model_residuals)

# ggplot(browseindex.model, aes(x=residuals))+
 # geom_histogram()

#Tried using ggplot but wasn't working out. The hist(model_residuals method worked though)
```

This histogram seems pretty normally distributed, with a bit of a right skew.

Now just trying autoplot which is not done in the source, but I think gives the same info that the author is coding for.

```{r}
autoplot(browseindex.model)
```

This isn't the greatest, but I don't think it's terrible. The residuals vs fitted line is a little bit off, and the normal q-q is also a bit off, but is straight for the most part. Scale-location has a big dip at the start, however. The residuals vs leverage seems ok, meaning there's not a sigificant outlier.

The source does a multicollinearity assumption check, but I'm not entirely sure what that is for or if it would be helpful for my analysis. I think it might be to see if there's a correlation between any of the predictor variables.

## Run the multiple regression

Use the anova function as done for linear regression with the model created.

```{r}
anova(browseindex.model)
```

From this, I can conclude that elevation and harvest intensity have a significant relationship with the moose browse index (p \<\< 0.01 for both).

The `summary()` function gives more information:

```{r}
summary(browseindex.model)
```

### Interpret

From this I can conclude that there is a significant relationship between moose browse index (the intercept) and elevation, as well as harvest intensity (F = 7.244, df = 6, 63, p \<0.01). The other predictors do not have a significant relationship with browse index.

-   The adjusted R squared tells that 35.19% of the variation can be explained by the predictors.

-   The first df is 6 because there are 7 variables used, so 7-1=6.

-   The second df is 63 because 86 observations weren't used because they weren't complete cases. 156-86=70, and 70-7=63.

## Adding in the binary data to see if it works

### Model

```{r}
browseindex.model2 <- lm(browse_index ~ elev_m + dist_h2o_m + dist_conifer_m + harvest_intensity_m2.ha + dist_dev_m + deer_density + harvest + hilltop + swamp + wetland_forest, data = Plot)
```

### Check assumptions

```{r}
autoplot(browseindex.model2)
```

This is similar to the run with just ratio data in all the plot, the residuals vs leverage is just different on the x-axis.

### Run model

```{r}
anova(browseindex.model2)
```

This shows the same predictors as being significant as in the first run without the binary data. It can be concluded that elevation and harvest intensity have a siginficant relationship with browse index (p\<\< 0.01)

```{r}
summary(browseindex.model2)
```

These results are also similar to the ratio predictors only run, however the significance is less. I can still conclude that elevation and harvest intensity have a significant relationship with browse index (F = 4.54, df = 10 and 59, p\<0.01 - although it's p\<0.05 for elevation).

-   The adjusted R-sq says that 33.91% of the variation in the response can be explained by the predictors.

-   10 degrees of freedom comes from 11 variables minus 1.

-   59 degrees of freedom comes from the 86 deleted observations from 156 = 76. I would think that next you would subtract 10 from 76, but that's not the case with 59 degrees of freedom, so I'm not positive where this number comes from.
