---
title: "Making new column Interaction"
format: html
editor: visual
---

Goal of this document is to make a new column that combines the original binary variables browsed, stripped, and broken to just one column: Interaction. Interaction will be binary 0(no) and 1(yes).

Worked with Erika Dec 18 to figure this out.

```{r}
rm(list = ls())
library(tidyverse)
library(broom) #for tidying model results
library(data.table) #for getting confidence intervals
library(bestglm) #for best subsets with glm models
library(MASS) #for step AIC
library(glmulti) #for another approach to model selection
library(oddsratio) #converts log odds to odds ratios
library(visreg) #for visualizing regression output
library(here)
```

Read in

```{r}
Old_tree <- read.csv(here("Data", "Moose_Tree_Data.csv"))
```

```{r}
Old_tree <- mutate(Old_tree, Interaction = rowSums(cbind(browsed, stripped, broken), na.rm = T))
```

Now fix a couple errors: the 6.3 is a typo, so change to 1. The values of 2 show that there was multiple types of interaction, but I just care whether there was interaction, so change these to 1.

```{r}
fix6.3 <- which(Old_tree$Interaction == 6.3)

Old_tree$Interaction[fix6.3] <- 1
#Success
```

```{r}
fix2 <- which(Old_tree$Interaction == 2)

Old_tree$Interaction[fix2] <- 1
#Success
```

Lastly, I want to remove browsed, stripped, and broken, as I'm not using them for predictors in my logistic regression.

```{r}
Old_tree <- Old_tree[-c(5:7)]
```

Save the new dataset to use for the multiple logistic regression.

```{r}
write.csv(Old_tree, here("Data", "FinalUpdated_Moose_Tree_Data.csv"))
```
