---
title: "Final Logistic Regression (Single)"
format: html
editor: visual
---

# Logistic Regression

In this file I will go through the workflow of running a logistic regression using data from the file: FinalUpdated_Moose_Tree_Data.csv

## Set up

```{r}
rm(list = ls())
library(tidyverse)
library(ggfortify)
library(oddsratio) #converts log odds to odds ratios
library(visreg) #for visualizing regression output
library(here)
```

Read in the data:

```{r}
Tree <- read.csv(here("Data", "FinalUpdated_Moose_Tree_Data.csv"))
```

Fix data types

```{r}
Tree$species <- as.factor(Tree$species)
```

Just want complete cases for the regression

```{r}
Tree <- Tree[complete.cases(Tree), ]
```

Went from 3913 rows to 3397 rows.

## Variables

Response:

-   Interaction (was the tree browsed, stripped, and/or broken by a moose) — Binary coded as 0(no) and 1(yes)

Predictor:

-   dbh_cm (the tree diameter at breast height) — Ratio

## Hypotheses

Question: Does dbh_cm have an impact on moose interaction?

[Null Hypothesis:]{.underline} dbh_cm is not a significant predictor of Interaction by moose.

[Alternative Hypothesis:]{.underline} dbh_cm is a significant predictor of interaction by moose.

[Biological Justification:]{.underline} According to the Massachusetts Division of Fisheries & Wildlife, moose prefer new growth for food, so a smaller dbh might be a predictor of Interaction.

## Plots

Make a new column with interaction coded as yes and no to make the plots look better.

```{r}
yes <- which(Tree$Interaction == 1)
no <- which(Tree$Interaction == 0)

Tree$InteractionYN <- NA

Tree$InteractionYN[yes] <- "Yes"
Tree$InteractionYN[no] <- "No"

Tree$InteractionYN <- as.factor(Tree$InteractionYN)
```

Look at the response variable

```{r}
ggplot(Tree, aes(InteractionYN))+
  geom_histogram(stat="count")+ #Allow for the x axis to be categorical, not continuous
  theme_bw()+ #remove gray background
  xlab("Interaction") #change x axis title
```

This shows that there were much more trees that were not interacted with, than those that were interacted with.

**Interaction and DBH:**

```{r}
ggplot(Tree, aes(x = dbh_cm, y = InteractionYN))+
  geom_point()+
  theme_bw()+ 
  xlab("DBH (cm)")+ #change x axis title
  ylab("Interaction") #change y axis title
```

This seems to show that there are more interactions at a smaller DBH, as the yes interaction points are more dense on the left. My guess is that dbh_cm will be a significant predictor or moose interaction.

## Fit and run the model

Use generalized linear model in order to tell R that responses are binary.

```{r}
dbh.mod <- glm(Interaction ~ dbh_cm, family = "binomial", data = Tree) #Fit the model with dbh as a function of interaction using the binomial family and the tree data set. 
summary(dbh.mod) 
```

The coefficients in this output indicate the average change in the odds of interaction occuring with each increase in dbh by 1 unit. This shows that with a 1 unit increase in dbh, the log odds of moose interaction *decreases* by 0.032484.

Now calculate the odds ratio to see what that means.

```{r}
or_glm(
  data = Tree, 
  model= dbh.mod, 
  incr = list(dbh_cm = 1))
```

This shows that the oddsratio for dbh_cm = 0.968. This means that for every 1 unit of increase in dbh, the odds that a moose will interact with that tree is 0.968, or 0.968 times less likely.

## Replot and Interpretation

Plot the results of the odds ratio

```{r}
ggplot(Tree, aes(dbh_cm, Interaction))+
  geom_point()+
  stat_smooth(method = "glm", method.args = list(family = "binomial"), se = F)
```

Now make a plot based on probabilities using the `visreg` package:

```{r}
visreg(dbh.mod, "dbh_cm", 
       gg = T, #use ggplot vs. base R
       scale = "response")+
  labs(x = "Tree DBH (cm)",
       y = "Prob. of moose interaction")+ #Change the axes titles
  theme_bw()
```

These results show that as tree DHB increases by 1, the probability of moose interaction with that tree decreases by 0.968. This is not a large decrease, but is significant (z = -7.071, p = 1.54e-12) based on the generalized linear model run. This result supports my alternative hypothesis, and is supported biologically. Moose prefer new growth trees (Massachusetts Division of Fisheries & Wildlife, n.d.), so they interact more with trees of a smaller dbh when browsing.

### Citations

Massachusetts Division of Fisheries & Wildlife. (n.d.). Moose in Massachusetts. Retrieved December 17, 2024, from <https://www.mass.gov/doc/living-with-moose-fact-sheet/download>
